---
title: "NFL Player Career Longevity"
author: "Imogen, Adam and Nolan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
status <- read.csv("status.csv")

```

```{r}
library(dplyr)
library(tidyverse)
```


```{r}
all_seasons <- read.csv("NFL Player Stats(1922 - 2022).csv")
head(all_seasons)
seasons <- all_seasons[all_seasons$Season >= 2006,] #only have injury data from 2006 onwards
seasons <- seasons %>%
  select(Tm, Season, Player, Age, Pos, G, GS, AllTD, Pts, FGM, FGA)
head(seasons)
dim(seasons)
all_injuries <- read.csv("totalConcussion.csv") #injury and concussion by player and year
all_injuries <-  all_injuries %>% separate(Player, into = c("FirstName", "LastName"), sep = " ")
all_injuries <- all_injuries[all_injuries$Position != 'Position',]

#clean seasons
seasons <-  seasons %>% separate(Player, into = c("FirstName", "LastName"), sep = " ")
seasons$LastName <- gsub("\\+$", "", seasons$LastName)
seasons <-  seasons %>% mutate(AllTD = ifelse(is.na(AllTD),0, AllTD))


```


```{r Clean Position Col}
seasons <- seasons %>%
  mutate(Pos = ifelse(Pos %in% c("LCB", "CB", "RCB", "RCB/LCB", "DB/RCB", "RCB/LCB/DB", "LCB/RCB", "DB/LCB", "RCB/DB",
                                 "SS", "FS", "SS/FS", "LCB/FS", "DB/FS", "FS/SS", "SS/RLB", "S", "DB"), "DB", Pos))

seasons <- seasons %>%
  mutate(Pos = ifelse(Pos %in% c("MLB", "ROLB", "LLB/ROLB", "RLB", "RLB/MLB", "LOLB", "LLB", "LILB", "LB", "LILB/ROL", 
                                 "RILB", "LLB/RLB", "LB/LLB", "LLB/MLB", "LILB/RIL", "ROLB/LOL", "OLB", "MLB/RLB"), "LB", Pos))

seasons <- seasons %>%
  mutate(Pos = ifelse(Pos %in% c("LT/RT", "RT/LT", "LT", "T", "NT", "RT"), "T", Pos))

seasons <- seasons %>%
  mutate(Pos = ifelse(Pos %in% c("LDT/RDT", "LDT", "DT", "DT/LDT", "RDT", "DL"), "DT", Pos))

seasons <- seasons %>%
  mutate(Pos = ifelse(Pos %in% c("RDE/LDE", "RDE", "DE", "LDE/RDE", "LDE"), "DE", Pos))

seasons <- seasons %>%
  mutate(Pos = ifelse(Pos %in% c("LG", "G"), "G", Pos))

seasons <- seasons %>%
  mutate(Pos = ifelse(Pos %in% c("C/LG"), "C", Pos))

seasons <- seasons %>%
  mutate(Pos = ifelse(Pos %in% c("RB/TE", "FB", "DT/FB"), "FB", Pos))

seasons <- seasons %>%
  mutate(Pos = ifelse(Pos %in% c("QB", "WR/QB"), "QB", Pos))

all_injuries <- all_injuries %>%
  mutate(Position = ifelse(Position %in% c("CB", "DB", "S"), "DB", Position))

seasons <- seasons[seasons$Pos != "",]
all_injuries <- all_injuries[all_injuries$Position != "LS",]
all_injuries <- all_injuries[all_injuries$Position != "KR",]
all_injuries <- all_injuries[all_injuries$Position != "PR",]

sort(unique(all_injuries$Position))
sort(unique(seasons$Pos))


```

```{r Clean Team Col}
 "df8d21cb55cec0801ca51a6f5319bec49bd5909f"

seasons$Team <- seasons$Tm

InjuriesSeason <- inner_join(all_injuries, seasons, by = "Team")

class(all_injuries$Team)
patterns <- c("Seahawks","Steelers", "Cowboys", "Lions", "Browns",       
"Chiefs","Dolphins","Eagles", "Buccaneers", "Vikings", "Saints", "Packers",
"Colts", "Falcons", "Bengals","Raiders", "Broncos", "Ravens", "Commanders",
"Giants", "49ers", "Jets", "Jaguars", "Panthers", "Bears", "Texans", "Patriots",
"Rams", "Chargers", "Titans", "Bills", "Cardinals", "Football Team", "Redskins",
"Niners")
replacements <- c("SEA", "PIT", "DAL", "DET", "CLE", "KAN", "MIA", "PHI",
              "TAM", "MIN", "NOR", "GNB", "IND", "ATL", "CIN", "LVR",
              "DEN", "BAL", "WAS", "NYG", "SFO", "NYJ", "JAX", "CAR", "CHI",
              "HOU", "NWE", "LAR", "LAC", "TEN", "BUF", "ARI", "WAS", "WAS", "SFO")
for (i in 1:length(patterns)) {
  all_injuries$Team <- gsub(patterns[i], replacements[i], all_injuries$Team)
}

head(all_injuries)
```


```{r Make the PlayerIDs}
#make injury ID to merge
all_injuries$PlayerID <- paste0(paste0(paste0(substring(all_injuries$FirstName, 1, 1), all_injuries$LastName), all_injuries$Position), all_injuries$Season)

#make season ID to merge
seasons$PlayerID <- paste0(paste0(paste0(substring(seasons$FirstName, 1, 1), seasons$LastName), seasons$Pos), seasons$Season)

#make status ID to merge
status$PlayerID <- paste0(paste0(paste0(substring(status$FirstName, 1, 1), status$LastName), status$Pos), status$Season)
```


```{r}
#MERGE ON INJURIES AND SEASONS
injuries <-  all_injuries %>% select(PlayerID, TotalInjuryCount, ConcussionTotal)
seasons2010 <- seasons[seasons$Season >= 2010,] %>% select(-Tm)
joined <- left_join(seasons2010, injuries, by = 'PlayerID')
joined <- joined %>% mutate(ConcussionTotal = ifelse(is.na(ConcussionTotal),0, ConcussionTotal))
joined <- joined %>% mutate(TotalInjuryCount = ifelse(is.na(TotalInjuryCount),0, TotalInjuryCount))

head(joined)
```


```{r Make final dataset}
#Making cumulative cols and response var
final <- joined
final <- final %>%
  group_by(FirstName, LastName, Pos) %>%
  arrange(Season) %>%
  mutate(G = cumsum(G), GS = cumsum(GS), AllTD = cumsum(AllTD), Pts = cumsum(Pts),
         FGM = cumsum(FGM), FGA = cumsum((FGA)), ConcussionTotal = cumsum(ConcussionTotal), TotalInjuryCount = cumsum(TotalInjuryCount))


#create response
final <- final %>% group_by(FirstName, LastName, Pos) %>%
  arrange(Season)%>% mutate(YearsRemaining = max(Season) - Season)

head(final)

final_all <-  final 

#filter positions that have very few samples
too_few <- final_all %>% group_by(Pos) %>% count(Pos)%>% filter(n < 100) %>% select(Pos) %>% ungroup() %>% pull(Pos)

#remove low sample size pos
final_all <- final_all %>% filter(!(Pos %in% too_few))

#table of average career lengths
average_len_careers <- final_all %>% group_by(FirstName, LastName, Pos) %>%
  arrange(Season)%>% mutate(careerLen = max(Season) - min(Season)) %>% ungroup() %>% select(Pos, careerLen) %>% group_by(Pos) %>% mutate(avgLen = mean(careerLen)) %>% distinct(Pos, avgLen) %>% arrange(desc(avgLen)) 

average_len_careers$avgLen <- round(average_len_careers$avgLen, 2)



#drop players that moved around a lot
final[final$Team != '3TM',]
final[final$Team != '2TM',]


#look at individual
final[final$LastName == 'Folk',]


kickers_final <- final[final$Pos == 'K',]
final_1 <- final
final_1 <- final[final$Pos != 'K',] %>% select(-FGM, -FGA)


final$ThreeYrLeft <- ifelse(final$YearsRemaining >= 3, 1 ,0)

final_1$ThreeYrLeft <- ifelse(final_1$YearsRemaining >= 3, 1 ,0)
final_1 <- final_1 %>% select(-YearsRemaining)
```
ThreeYearsLeft is 1, if you have 3 or more years left in the NFL and 0 if you have less than 3 years

```{r Variable Exploration}
final %>% ungroup() %>%
  dplyr::select(YearsRemaining ,Age, G, GS, AllTD, Pts, TotalInjuryCount, ConcussionTotal) %>%
  pivot_longer(cols = -YearsRemaining) %>% 
  # Plot every value against whiffs
  ggplot(aes(value, YearsRemaining)) +
  geom_point() +
  
  geom_smooth(method = "lm") +
  # Put each variable into its own plot:
  facet_wrap(vars(name), scales = "free") + theme_minimal()

```
Age looks like it will be the best factor to predict years remaining as the trend line shows a marginal negative correlation, meaning as age increases the number of years remaining you have decreases.
```{r}
library(caret)

set.seed(123)
finalRB <- 
train_index <- createDataPartition(final$YearsRemaining, p = 0.8, list = FALSE)
train_data <- final[train_index, ]
test_data <- final[-train_index, ]

# Check the dimensions
dim(train_data)
dim(test_data)


# Ensure YearsRemaining is numeric
train_data$YearsRemaining <- as.numeric(as.character(train_data$YearsRemaining))
train_data$YearsRemaining <- as.numeric(as.character(train_data$YearsRemaining))

names(final)

train_dataRB <- train_data[train_data$Pos == 'RB',] 
train_dataRB<- train_dataRB %>% ungroup() %>% select(YearsRemaining,Age, G, GS, AllTD, Pts, TotalInjuryCount, ConcussionTotal)

test_dataRB <- test_data[test_data$Pos == 'RB',]
test_dataRB<- test_dataRB %>% ungroup() %>% select(Season, YearsRemaining, Age, G, GS, AllTD, Pts, TotalInjuryCount, ConcussionTotal)

linearRB <- lm(YearsRemaining ~ ., train_dataRB)
# 
# 
# # Get predictions with 95% confidence intervals
# predictions <- predict(linearRB, newdata = test_dataRB)
# intervals <- predictions
# intervals$lwr <- predictions - 2
# intervals$upp <- predictions + 2
# intervals_1 <- ifelse(test_dataRB$YearsRemaining >= intervals$lwr & test_dataRB$YearsRemaining <= intervals$upp, 1, 0)

summary(linearRB)
```

The significant factors at 5% were Age, Games, Touchdowns, Points and Concussion Total. These will be the features to use in random effects.

```{r RB Logistic }
library(caret)

set.seed(123)

finalRB <- final_1[final_1$Pos == 'RB',]

train_index <- createDataPartition(finalRB$ThreeYrLeft, p = 0.8, list = FALSE)
train_dataRB <- finalRB[train_index, ]
test_dataRB <- finalRB[-train_index, ]

# Check the dimensions
dim(train_dataRB)
dim(test_dataRB)


# Ensure as.factor(train_data$FiveYrLeft)
train_dataRB$ThreeYrLeft <- as.numeric(train_dataRB$ThreeYrLeft)
test_dataRB$ThreeYrLeft <- as.numeric(test_dataRB$ThreeYrLeft)

names(final)

train_dataRB<- train_dataRB %>% ungroup() %>% select(ThreeYrLeft,Age, G, GS, AllTD, Pts, TotalInjuryCount, ConcussionTotal)

test_dataRB<- test_dataRB %>% ungroup() %>% select(ThreeYrLeft, Age, G, GS, AllTD, Pts, TotalInjuryCount, ConcussionTotal)

logisticRB <- glm(ThreeYrLeft ~ ., train_dataRB, family = "binomial")
summary(logisticRB)

sjPlot::plot_model(logisticRB, type ="pred") #part of logistic regression curve

#AIC: 1329.6
```
Our first model is a linear model for RB using all features and of those features Age, Pts Scored, TD Scored and Games Played were significant. This model had an AIC of 1330. Next we ran a mixed effects linear model using the significant factors.


```{r RB Mixed Effects}

logisticRB1 <- glm(ThreeYrLeft ~ Age +  G * Pts * AllTD, train_dataRB, family = "binomial")
summary(logisticRB1)
#AIC: 1286

predictions <- predict(logisticRB1, newdata = test_dataRB, type = "response")

```
In the interaction model, only two variables appeared significant at the 5% level. These were Games Played and THe interaction between Age and Games Played.


```{r}
#Threshold Testing
curr = 0
for (i in seq(0.3, 0.7, by = 0.01)){
  test_pred <- ifelse(is.na(predictions), NA, ifelse(predictions > i, 1, 0))
  
  accuracy <- mean(test_pred == test_dataRB$ThreeYrLeft)
  
  
  if (accuracy > curr){
    print(accuracy)
    best = c(i, accuracy)
    curr = accuracy
  }
 
}

test_pred <- ifelse(is.na(predictions), NA, ifelse(predictions > best[1], 1, 0))
conf_matrix <- table(test_pred, test_dataRB$ThreeYrLeft)


# Calculate metrics
precision <- conf_matrix[2, 2] / sum(conf_matrix[2, ])
recall <- conf_matrix[2, 2] / sum(conf_matrix[, 2])
f1_score <- 2 * (precision * recall) / (precision + recall)
accuracy <- mean(test_pred == test_dataRB$ThreeYrLeft)

```
Our accuracy for this model was 70% meaning that 70/100 samples predicted correctly whether this athlete will retire in the next 3 years or not

WR Accuracy Model
```{r WR Logistic }

finalWR <- final_1[final_1$Pos == 'WR',]

train_index <- createDataPartition(finalWR$ThreeYrLeft, p = 0.8, list = FALSE)
train_dataWR <- finalWR[train_index, ]
test_dataWR <- finalWR[-train_index, ]

# Check the dimensions
dim(train_dataWR)
dim(test_dataWR)


# Ensure as.factor(train_data$FiveYrLeft)
train_dataWR$ThreeYrLeft <- as.numeric(train_dataWR$ThreeYrLeft)
test_dataWR$ThreeYrLeft <- as.numeric(test_dataWR$ThreeYrLeft)

train_dataWR<- train_dataWR %>% ungroup() %>% select(ThreeYrLeft,Age, G, GS, AllTD, Pts, TotalInjuryCount, ConcussionTotal)

test_dataWR <- test_dataWR %>% ungroup() %>% select(ThreeYrLeft, Age, G, GS, AllTD, Pts, TotalInjuryCount, ConcussionTotal)

logisticWR <- glm(ThreeYrLeft ~ ., train_dataWR, family = "binomial")
summary(logisticWR)

#AIC of 2034.7

sjPlot::plot_model(logisticWR, type ="pred")
```
The plot is the same concept as the prediction model we ran in line 271, just adapted to wide receivers.

Of the features used in our WR logistic regression Age, Games Played, Games Started, TD Scored and Points Scored were all significant at the 5% level so we will used these in a mixed effects model.

```{r WR Mixed Effects}
logisticWR1 <- glm(ThreeYrLeft ~ Age * Pts * AllTD * G * GS, train_dataWR, family = "binomial")
summary(logisticWR1)
#AIC: 1947.4
 
predictions <- predict(logisticWR1, newdata = test_dataWR, type = "response")
```
Age and the interaction between Age, Points, Touchdowns and Games were significant at 5%.

```{r}

#Threshold Testing
curr = 0
for (i in seq(0.3, 0.7, by = 0.01)){
  test_pred <- ifelse(is.na(predictions), NA, ifelse(predictions > i, 1, 0))
  
  accuracy <- mean(test_pred == test_dataWR$ThreeYrLeft)
  
  
  if (accuracy > curr){
    best = c(i, accuracy)
    curr = accuracy
  }
  
}

test_pred <- ifelse(is.na(predictions), NA, ifelse(predictions > best[1], 1, 0))
conf_matrix <- table(test_pred, test_dataWR$ThreeYrLeft)

# Calculate metrics
precision <- conf_matrix[2, 2] / sum(conf_matrix[2, ])
recall <- conf_matrix[2, 2] / sum(conf_matrix[, 2])
f1_score <- 2 * (precision * recall) / (precision + recall)
accuracy <- mean(test_pred == test_dataWR$ThreeYrLeft)
accuracy
```
After threshold testing, our best model was 67% accurate meaning 67/100 samples were predicted correctly on whether they will retire in the next 3 years.


```{r Random Effects}
library(lme4)
pos_data <- final_all %>% ungroup() %>% select(Age, Pos, G, GS, AllTD, Pts, TotalInjuryCount, ConcussionTotal, YearsRemaining)




position_age <- lmer(YearsRemaining ~ Age + (1|Pos), data = pos_data)
summary(position_age)
random_effects = ranef(position_age)
ranef_df <-  data.frame(
  int = random_effects$Pos$`(Intercept)`,
  pos = rownames(random_effects$Pos)
)
performance::r2(position_age)
ggplot(ranef_df, aes(x=int, y=reorder(pos, int))) +
  geom_point() + 
  labs(x= "intercept", y="pos", title = 'Effect of Position on Years Remaining', subtitle = 'lm ~ Age') +
  theme_minimal()
```
Age: for every 0.15 years older, 1 year decrease in number of years remaining

```{r Games}
scaled <- data.frame(cbind(scale(pos_data[,c('Age', 'G', 'YearsRemaining')]), pos_data$Pos)) %>% rename( 'Pos'='V4')
scaled$Age <- as.numeric(scaled$Age)
scaled$YearsRemaining <- as.numeric(scaled$YearsRemaining)
scaled$G <- as.numeric(scaled$G)



position_g <- lmer(YearsRemaining ~ G * Age + (1|Pos), data = scaled)
summary(position_g)
random_effects = ranef(position_g)
ranef_df <-  data.frame(
  int = random_effects$Pos$`(Intercept)`,
  pos = rownames(random_effects$Pos)
)
performance::r2(position_g)
ggplot(ranef_df, aes(x=int, y=reorder(pos, int))) +
  geom_point() + 
  labs(x= "intercept", y="pos", title = 'Effect of Position on Years Remaining', subtitle = 'lm ~ Age * G') +
  theme_minimal()
```
Age is the greatest contributor to this model as it has highest t value. From the model you can see a negative correlation between age and years remaining. As age increases by 1 year, number of years remaining decreases by 0.23 years.

Number of games has a relatively small impact on the model. It has a small positive correlation with years remaining, meaning for every game played, there is a 0.02 increase in years remaining. Depending on how you interpret the model this makes sense as if you are a better/more active players, maybe you are more likely to keep playing vs someone who doesn't get a lot of reps and retires early.

The interaction between Games and Age has a relatively small impact on the model. It has a small negative correlation meaning that for every 1 year increase in age, the effect of number of games played on years remaining decreases by 0.028.

```{r Compare to average career lengths}
library(DT)
datatable(average_len_careers)

```

